Nested sampling with two objective functions

### 2015-11-02 Paper/ms.tex
Dnest
Random work

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{We present a generalization of the Nested Sampling algorithm to a
broad class of problems which arise in statistical mechanics.}

@@

Nested Sampling \citep[NS][]{skilling} is an effective and popular
Monte Carlo algorithm which can solve a wide range of problems in Bayesian
computation and statistical mechanics
\citep{2009arXiv0906.3544P, 2014PhRvE..89b2302P, 2015arXiv150303404B}.
In a Bayesian inference problem with unknown parameters denoted collectively
by a vector $\xx$, the
posterior distribution for the parameters given data $\data$ (and prior
information $\II$) is:
\begin{eqnarray}
p(\xx | \data, \II) &=&
\frac{p(\xx | \II)p(\data | \xx, \II)}{p(\data | \II)}\\
&=& \frac{\pi(\xx)L(\xx)}{Z}
\end{eqnarray}
where $\pi(\xx)$ is the prior distribution, $L(\xx)$ is the likelihood
function, and $Z$ is the normalising constant, usually known as the
``marginal likelihood'' or the ``evidence'':
\begin{eqnarray}
Z &=& \int \pi(\xx) L(\xx) \, \dx.\label{eqn:evidence}
\end{eqnarray}
NS begins by drawing particles from the
prior $\pi(\xx)$ and successively imposing constraints on the value of
the likelihood $L(\xx)$ that compress the prior mass by an
approximately-known factor. Like other related algorithms such as those
based on ``annealing'', NS
moves through a sequence of probability distributions beginning from the
prior. This sequence of distributions is defined by
\begin{eqnarray}
p(\xx; \ell) &=& \frac{1}{X(\ell)}\pi(\xx)\mathds{1}\left(L(\xx) > \ell\right).
\end{eqnarray}

This enables the calculation of the marginal likelihood $Z$
and properties of the posterior $\pi(\xx)L(\xx)/Z$.
NS is required because Equation~\ref{eqn:evidence}
is the expected value of $L$ with respect to a very heavy-tailed distribution
(the prior for the $L$ value implied by $\pi(\xx)$). Hence, there is a strong
connection between NS and ideas from rare event simulation \citep{walter}.

NS also allows us to calculate the properties of any other distribution that
is in some sense intermediate
between the prior and the posterior. For example, we might be interested in
a ``power posterior'' where the likelihood is raised to a power $\beta$:
\begin{eqnarray}
p(\xx; \beta) &=& \frac{\pi(\xx)L(\xx)^\beta}{Z(\beta)}\label{eqn:power_posterior}
\end{eqnarray}
The normalisation and posterior samples from this distribution can be obtained
from the original NS run by re-weighting the output according to $L(\xx)^\beta$
instead of the usual $L(\xx)$ used to obtain the posterior.
An example application, in the case of
a Bayesian inference problem with a ``gaussian noise'' likelihood assumption,
computing $p(\xx; \beta)$ for $\beta \neq 1$ allows us to explore what the
posterior distribution would have been if the noise variance had been greater.
This is different from including an extra parameter to allow the noise variance
to be greater, because NS allows you to test the consequences of values of the
variance that are very implausible given the data.

@@

In some inference and
statistical mechanics problems, there are two or more scalar functions of
$\xx$ that are relevant. Suppose our prior is $\pi(\xx)$ as before, and
we obtain testable information that fixes the expected values of two scalar
functions of $\xx$, $L_1(\xx)$ and $L_2(\xx)$. It is well known that the
updated probability distribution that takes into account the constraints is
of the ``canonical'' form:

\begin{eqnarray}
p(\xx; \beta_1, \beta_2) &=& \frac{\pi(\xx)\exp\left[\beta_1L_1(\xx)+\beta_2L_2(\xx)\right]}
{Z(\beta_1, \beta_2)}
\end{eqnarray}

If we were only interested in a single canonical distribution, for example
with $\beta_1 = 0.3$ and $\beta_2 = 0.7$, we could estimate its normalising
constant and by running standard Nested Sampling with ``likelihood''
$L(\xx) = \exp\left[0.3L_1(\xx) + 0.7L_2(\xx)\right]$. However, usually we
are interested in a range of values for $\beta_1$ and $\beta_2$, and we
want to know the normalisation $Z(\beta_1, \beta_2)$, called the
partition function. This work describes
progress towards solving this class of problems while maintaining the benefits
of Nested Sampling, such as the ability to cope with first-order phase
transitions. The algorithm is designed for problems with two objective
functions, but should generalize to problems with more than two (though perhaps
not more than a few).

The prior $\pi(\xx)$ implies a certain prior for $L_1$ and $L_2$, which we
denote $\pi(L_1, L_2)$. An example of a prior is shown as the
probability density in Figure~\ref{fig:joint1}. The partition function
$Z(\beta_1, \beta_2)$ is a set of expected values with respect to this density.
Simple Monte Carlo sampling from $\pi$ will not work except for values of
$(\beta_1, \beta_2)$ where the canonical distribution remains similar to $\pi$.
We need a sampler that explores regions where $L_1$ and $L_2$ are high in order
to accurately estimate $Z(\beta_1, \beta_2)$.

\begin{figure}
\centering
\includegraphics[scale=0.5]{figures/joint1.pdf}
\caption{\it An example of a prior for $L_1$ and $L_2$, which is implied by
the prior $\pi(\xx)$. We have overplotted some points drawn from this prior.
\label{fig:joint1}}
\end{figure}

Our proposed algorithm (introduced in Section~\ref{sec:algorithm}) makes use
of a quantity we call the corner count. For any values $(\ell_1, \ell_2)$,
the prior mass for which both objective functions are lower is
\begin{eqnarray}
X(\ell_1, \ell_2) &=& \int \pi(L_1, L_2) \mathds{1}\left(L_1 < \ell_1, L_2 < \ell_2 \right)
 \, dL_1 \, dL_2
\end{eqnarray}
This is the two dimensional cumulative distribution function (CDF) of
$L_1$ and $L_2$.

The corner count of particle $i$ is defined as the number of particles with
a lower value of {\em both} objective functions compared to particle $i$.
Graphically, on a plot of $L_2$ vs. $L_1$ the corner count is the number of
particles in the rectangle whose upper right corner is at
$\left(L_1(\xx_i), L_2(\xx_i)\right)$ (see Figure~\ref{fig:joint2} for an
example). The equation for the corner count of particle $i$ is:
\begin{eqnarray}
C_i &=& \sum_{\substack{j=1\\{j \neq i}}}^N \mathds{1}\left(L_1(\xx_j) < L_1(\xx_i),
L_2(\xx_j) < L_2(\xx_i)
\right).\label{eqn:corner_count}
\end{eqnarray}

@@

>>>>

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{We present a generalization of the Nested Sampling algorithm for
estimating the partition function of a two-dimensional family of
probability distributions.}

@@

Nested Sampling \citep[NS][]{skilling} is an effective and popular
Monte Carlo algorithm which can solve a wide range of problems in Bayesian
computation and statistical mechanics
\citep{2009arXiv0906.3544P, 2014PhRvE..89b2302P, 2015arXiv150303404B}.
In a Bayesian inference problem with unknown parameters denoted collectively
by a vector $\xx$, the
posterior distribution for the parameters given data $\data$ (and prior
information $\II$) is:
\begin{eqnarray}
p(\xx | \data, \II) &=&
\frac{p(\xx | \II)p(\data | \xx, \II)}{p(\data | \II)}\\
&=& \frac{\pi(\xx)L(\xx)}{Z}
\end{eqnarray}
where $\pi(\xx)$ is the prior distribution, $L(\xx)$ is the likelihood
function, and $Z$ is the normalising constant, usually known as the
``marginal likelihood'' or the ``evidence'':
\begin{eqnarray}
Z &=& \int \pi(\xx) L(\xx) \, \dx.\label{eqn:evidence}
\end{eqnarray}
The main goal of NS is to approximate $Z$.
This is done by drawing particles from the
prior $\pi(\xx)$ and successively imposing constraints on the value of
the likelihood $L(\xx)$ that compress the prior mass by an
approximately-known factor. Like other related algorithms such as those
based on ``annealing'', NS
moves through a sequence of probability distributions beginning from the
prior. This sequence of distributions is defined by
\begin{eqnarray}
p(\xx; \ell) &=& \frac{1}{X(\ell)}\pi(\xx)\mathds{1}\left(L(\xx) > \ell\right).
\end{eqnarray}

In addition to the marginal likelihood $Z$, NS also allows us
to approximate properties of the posterior $\pi(\xx)L(\xx)/Z$.
NS is required because Equation~\ref{eqn:evidence}
is the expected value of $L$ with respect to a very heavy-tailed distribution
(the prior for the $L$ value implied by $\pi(\xx)$). Hence, there is a strong
connection between NS and ideas from rare event simulation \citep{walter}.
NS also allows us to calculate the properties of any other distribution that
is in some sense intermediate
between the prior and the posterior. For example, we might be interested in
a ``power posterior'' where the likelihood is raised to a power $\beta$:
\begin{eqnarray}
p(\xx; \beta) &=& \frac{\pi(\xx)L(\xx)^\beta}{Z(\beta)}\label{eqn:power_posterior}
\end{eqnarray}
The normalisation and posterior samples from this distribution can be obtained
from the original NS run by re-weighting the output according to $L(\xx)^\beta$
instead of the usual $L(\xx)$ used to obtain the posterior.
An example application, in the case of
a Bayesian inference problem with a ``gaussian noise'' likelihood assumption,
computing $p(\xx; \beta)$ for $\beta \neq 1$ allows us to explore what the
posterior distribution would have been if the noise variance had been greater.
This is different from including an extra parameter to allow the noise variance
to be greater, because NS allows you to test the consequences of values of the
variance that are very implausible given the data.

@@

In some inference and
statistical mechanics problems, there are two or more scalar functions of
$\xx$ that are relevant. Suppose our prior is $\pi(\xx)$ as before, and
we obtain testable information that fixes the expected values of two scalar
functions of $\xx$, $L_1(\xx)$ and $L_2(\xx)$. It is well known that the
updated probability distribution that takes into account the constraints is
of the ``canonical'' form:
\begin{eqnarray}
p(\xx; \beta_1, \beta_2) &=& \frac{\pi(\xx)\exp\left[\beta_1L_1(\xx)+\beta_2L_2(\xx)\right]}
{Z(\beta_1, \beta_2)}
\end{eqnarray}
If we were only interested in a single canonical distribution, for example
with $\beta_1 = 0.3$ and $\beta_2 = 0.7$, we could estimate its normalising
constant and by running standard Nested Sampling with ``likelihood''
$L(\xx) = \exp\left[0.3L_1(\xx) + 0.7L_2(\xx)\right]$. However, usually we
are interested in a range of values for $\beta_1$ and $\beta_2$, and we
want to know the normalisation $Z(\beta_1, \beta_2)$, called the
partition function. This work describes
progress towards solving this class of problems while maintaining the benefits
of Nested Sampling, such as the ability to cope with first-order phase
transitions. The algorithm is designed for problems with two objective
functions, but should generalize to problems with more than two (though perhaps
not more than a few).

The prior $\pi(\xx)$ implies a certain prior for $L_1$ and $L_2$, which we
denote $\pi(L_1, L_2)$. An example of a prior is shown as the
probability density in Figure~\ref{fig:joint1}. The partition function
$Z(\beta_1, \beta_2)$ is a set of expected values with respect to this density.
Simple Monte Carlo sampling from $\pi$ will not work except for values of
$(\beta_1, \beta_2)$ where the canonical distribution remains similar to $\pi$.
We need a sampler that explores regions where $L_1$ and $L_2$ are high in order
to accurately estimate $Z(\beta_1, \beta_2)$.
\begin{figure}
\centering
\includegraphics[scale=0.5]{figures/joint1.pdf}
\caption{\it An example of a prior for $L_1$ and $L_2$, which is implied by
the prior $\pi(\xx)$. We have overplotted some points drawn from this prior.
\label{fig:joint1}}
\end{figure}

Our proposed algorithm (introduced in Section~\ref{sec:algorithm}) makes use
of a quantity we call the corner count. For any values $(\ell_1, \ell_2)$,
the prior mass for which both objective functions are lower is
\begin{eqnarray}
X(\ell_1, \ell_2) &=& \int \pi(L_1, L_2) \mathds{1}\left(L_1 < \ell_1, L_2 < \ell_2 \right)
 \, dL_1 \, dL_2
\end{eqnarray}
This is the two dimensional cumulative distribution function (CDF) of
$L_1$ and $L_2$. The corner count is a related quantity when we have
a collection of $N$ particles in the $L_1$-$L_2$ plane.
The corner count of particle $i$ is defined as the number of particles with
a lower value of {\em both} objective functions compared to particle $i$:
\begin{eqnarray}
C_i &=& \sum_{\substack{j=1\\{j \neq i}}}^N \mathds{1}\left(L_1(\xx_j) < L_1(\xx_i),
L_2(\xx_j) < L_2(\xx_i)
\right).\label{eqn:corner_count}
\end{eqnarray}
Graphically, on a plot of $L_2$ vs. $L_1$ the corner count is the number of
particles in the rectangle whose upper right corner is at
$\left(L_1(\xx_i), L_2(\xx_i)\right)$ (see Figure~\ref{fig:joint2} for an
example).

@@

\bibitem[\protect\citeauthoryear{Brewer, P{\'a}rtay,
\& Cs{\'a}nyi}{2011}]{dnest} Brewer B.~J., P{\'a}rtay L.~B., Cs{\'a}nyi G., 2011,
Statistics and Computing, 21, 4, 649-656. arXiv:0912.2380
